Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-03-19 Manuscript/HI_Model_ms.Rmd
updated manuscript files

### *4.3 Limitations of this study* 
There is ample evidence that increasing atmospheric CO~2~ concentrations can stimulate NPP [@norby_forest_2005; @zhu_greening_2016], but the magnitude and persistence of this effect remains highly uncertain [@franks_sensitivity_2013; @walker_integrating_2020]. Our results demonstrate that long-term projections of ecosystem carbon balance are highly sensitive to uncertainty in CFE strength. With no CFE, Hawaiian ecosystems transitioned into a net source of CO~2~ to the atmosphere beginning in the latter half of the 21st century under the high land use and high radiative forcing scenario. However, a 5% increase in NPP for every 100 ppm increase in atmospheric CO~2~ applied to the same scenario resulted in Hawaiian ecosystems remaining a net carbon sink throughout the entire simulation period. A 15% CFE applied to the high land use and high radiative forcing scenario resulted in a nearly 5-fold increase in mean annual NBP averaged across all years and Monte Carlo realizations. Despite this demonstrated sensitivity to a CFE, the combined effect of several potentially attenuating factors makes it difficult to determine the appropriate magnitude of a CFE with any degree of confidence [@walker_integrating_2020]. Nitrogen and phosphorus limitation can reduce or eliminate a CFE [@reich_nitrogen_2006; @norby_co2_2010; @he_nitrogen_2017; @terrer_nitrogen_2019], as can water limitation and heat stress [@obermeier_reduced_2017; @birami_hot_2020]. Forest age may also be a factor, with young aggrading forests showing a strong positive growth response to CO~2~ fertilization [@walker_decadal_2019], while old-growth forests show little to no response [@jiang_fate_2020; @yang_low_2020]. This evidence indicates that a CFE may be highly variable across space and time, suggesting that it may be unrealistic to apply a single CFE rate value across an entire region over several decades. Until mechanistic understanding is improved, the most conservative approach when projecting future ecosystem balance in the context of climate mitigation planning may be to assume no CFE, with the knowledge that any realized CFE will only enhance ecosystem carbon sequestration.

Our model does not currently differentiate between forests dominated by native versus non-native tree species, which could influence estimates of ecosystem carbon balance. Native forests in the Hawaiian Islands are dominated by ʻōhiʻa (_Metrosideros polymorpha_ Gauditch), an endemic foundational tree species found across a broad range of climatic and edaphic conditions [@ziegler_hawaiian_2002]. Beginning in 2010, a fungal disease termed Rapid ʻŌhiʻa Death (ROD) caused by two _Ceratocystus_ spp. has emerged and caused widespread mortality to mature ʻōhiʻa trees across a range of size classes, primarily on Hawaiʻi Island [@mortenson_assessing_2016; @fortini_evolving_2019]. The distribution and potential range of this emerging threat to Hawaiʻi's most widespread native tree species has only recently been mapped [@vaughn_approach_2018; @fortini_evolving_2019], but the pulse of newly dead organic matter and reduction in photosynthetic capacity induced by widespread tree mortality could significantly alter ecosystem carbon balance over the long-term [@sleeter_effects_2019]. ROD-affected forest canopies could also potentially see a replacement of long-lived, stress-tolerating ʻōhiʻa with non-native tree species that have faster growth rates, lower wood density, and faster tissue turnover, all of which could influence the long-term trajectory of forest carbon cycling in Hawaiʻi. New model projections for Hawaiʻi that incorporate ROD spread rates and forest restoration scenarios will therefore require differentiation among forest ecosystems dominated by native and non-native tree species.

>>>>

### *4.3 Limitations of this study* 
There is ample evidence that increasing atmospheric CO~2~ concentrations can stimulate NPP [@norby_forest_2005; @zhu_greening_2016], but the magnitude and persistence of this effect remains highly uncertain [@franks_sensitivity_2013; @walker_integrating_2020]. Our results demonstrate that long-term projections of ecosystem carbon balance are highly sensitive to uncertainty in CFE strength. With no CFE, Hawaiian ecosystems became a net source of CO~2~ to the atmosphere beginning in the latter half of the 21st century under the high land use and high radiative forcing scenario. However, a CFE equivalent to a 5% increase in NPP for every 100 ppm increase in atmospheric CO~2~ applied to the same scenario resulted in Hawaiian ecosystems remaining a net carbon sink throughout the entire simulation period. A 15% CFE applied to the high land use and high radiative forcing scenario resulted in a nearly 5-fold increase in mean annual NBP averaged across all years and Monte Carlo realizations. Despite this demonstrated sensitivity to a CFE, several potentially attenuating factors make complicate selecting a realistic CFE value with any degree of confidence [@walker_integrating_2020]. Nitrogen and phosphorus limitation can reduce or eliminate a CFE [@reich_nitrogen_2006; @norby_co2_2010; @he_nitrogen_2017; @terrer_nitrogen_2019], as can water limitation and heat stress [@obermeier_reduced_2017; @birami_hot_2020]. Forest age may also be a factor, with young aggrading forests showing a strong positive growth response to CO~2~ fertilization [@walker_decadal_2019], while old-growth forests show little to no response [@jiang_fate_2020; @yang_low_2020]. This evidence indicates that a CFE may be highly variable across space and time, suggesting that it may be unrealistic to apply a single CFE rate value across an entire region over several decades. Until mechanistic understanding is improved, the most conservative approach when projecting future ecosystem balance in the context of climate mitigation planning may be to assume no CFE, with the knowledge that any realized CFE will only enhance ecosystem carbon sequestration.

Our model does not currently differentiate between forests dominated by native versus non-native tree species, which could influence estimates of ecosystem carbon balance. Native forests in the Hawaiian Islands are dominated by ʻōhiʻa (_Metrosideros polymorpha_ Gauditch), an endemic foundational tree species found across a broad range of climatic and edaphic conditions [@ziegler_hawaiian_2002]. Beginning in 2010, a fungal disease termed Rapid ʻŌhiʻa Death (ROD) caused by two _Ceratocystus_ spp. has emerged and caused widespread mortality to mature ʻōhiʻa trees across a range of size classes, primarily on Hawaiʻi Island [@mortenson_assessing_2016; @fortini_evolving_2019]. The distribution and potential range of this emerging threat to Hawaiʻi's dominant native tree species have only recently been mapped [@vaughn_approach_2018; @fortini_evolving_2019], but the pulse of newly dead organic matter and reduction in photosynthetic capacity induced by widespread tree mortality could significantly alter ecosystem carbon balance over the long-term [@sleeter_effects_2019].  ROD-affected forests could also undergo a replacement of canopy dominant stress-tolerating ʻōhiʻa trees by non-native tree species with faster relative growth rates, lower wood density, and faster tissue turnover, which could alter the long-term trajectory of forest carbon cycling in Hawaiʻi. New model projections for Hawaiʻi that incorporate ROD spread rates and forest restoration scenarios will therefore require differentiation among forest ecosystems dominated by native and non-native tree species.

Interannual climate variability is a primary factor influencing spatial and temporal patterns in global wildland fire activity [@abatzoglou_global_2018], and climate warming is expected to increase wildland fire frequency and wildfire season length across a wide range of biomes [@sun_global_2019]. Although our model projections capture the spatial and temporal variation in ignition probability and fire extent by sampling from previous fire years (1999-2019), we did not incorporate how projected future climate change under different RCPs will affect the frequency and extent of wildland fire. Recent fire probability modeling for the northwest portion of Hawaiʻi Island indicated that projected drying and warming trends under RCP 8.5 could increase maximum fire probability values more than three-fold and shift areas of peak flammability to higher elevation by mid-century [@trauernicht_vegetationrainfall_2019]. Extending this probability fire modeling approach statewide would provide a quantitative, spatially explicit assessment of wildland fire probability for the main Hawaiian Islands as predicted by climate, land cover, and ignition density, which is highly correlated with population density [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019]. This approach would provide future simulation model projections of ecosystem carbon balance with more realistic scenarios of expected annual area burned based on the integrated effects of future climate and land use change.
